---
title: "wandering_morphemes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# install CRAN packages (if not yet installed)
sapply(c("data.table", "tidyverse", "devtools", "party", "Hmisc", "rms", "RColorBrewer"), function(x) 
  if(!is.element(x, installed.packages())) install.packages(x, dependencies = T))
# install non-CRAN packages (if not yet installed)
if(!is.element("wizard", installed.packages())) {
  devtools::install_github("hartmast/wizard")
}

# load packages
library(tidyverse)
library(data.table)
library(wizard)
library(party)
library(Hmisc)
library(rms)
library(RColorBrewer)

# quick barplot function
qbarplot2 <- function(df, ..., color = c("Blues", F), angle = 0) {
  
  if(missing(color)) { color = "Blues"  }
  
  p <- qbarplot(df, cex = 5, ...) + theme_bw() +
    theme(axis.text = element_text(size = 18)) +
    theme(axis.title = element_text(size = 18)) +
    theme(strip.text = element_text(size = 18)) +
    theme(legend.text = element_text(size = 18)) +
    theme(legend.title = element_text(size = 18, face = "bold")) +
    theme(text = element_text(size = 18))
  
  if(angle!=0) {
    p <- p + theme(axis.text.y = element_text(size = 18)) +
      theme(axis.text.x = element_text(angle = angle, size = 18,hjust = .9))
  } else {
    if(angle!=0) {
     p <- p + theme(axis.text = element_text(size = 18))
    }
  }
  
  
  if(color == "Blues") {
    p <- p + scale_fill_brewer("Blues")
  }
  
  return(p)
    
}

```

## Session info (for reproducability)

```{r seesioninfo}

sessionInfo()

```

## Read data

This code reads and binds the data:

```{r data, message = FALSE, warning = FALSE, echo = TRUE, results = 'hide'}

# read data
ws <- read_csv("wegens/wegens_DECOW16BX_anno.csv")
ws <- mutate(ws, cxn = "wegens")
w0 <- read_csv("wegens/wegen_sample_DECOW16BX_anno.csv")
w0 <- mutate(w0, cxn = "wegen")
uws <- read_csv("willens/um_willens_DECOW16B.csv")
uws <- mutate(uws, cxn = "willens")
uw0 <- read_csv("willens/um_willen_DECOW16B_sample.csv")
uw0 <- mutate(uw0, cxn = "willen")

# add empty position column for willens (so that rbind works in the next step)
uws <- mutate(uws, position = NA)
uw0 <- mutate(uw0, position = NA)

# bind all together
w <- rbind(uws, uw0, ws, w0)



```


## Gender distribution

This code plots the distribution of grammatical genders in the data:

```{r genders, out.width = '50%', warning = FALSE}

# distinction between strong and weak masculines
w$class <- NA
w[which(w$Gender %in% c("m", "n") & w$s_less=="x" & w$Number=="sg"),]$class <- "weak"
w[which(is.na(w$class)),]$class <- "not_weak"

# willen(s)
w %>% filter(cxn %in% c("willens", "willen")) %>%
  filter(Gender %in% c("f", "m", "n")) %>%
  qbarplot2(cxn, Gender,
                                                            ylab = "Relative Frequency", xlab = "Variant") + guides(fill = guide_legend(title = "Gender"))


# for types rather than tokens
willen_s <- filter(w, cxn %in% c("willens", "willen"))
willen_s[!duplicated(willen_s$lemma),] %>%
    filter(Gender %in% c("f", "m", "n")) %>%
  qbarplot2(cxn, Gender,
                                                            ylab = "Relative Frequenz (Types)", xlab = "Variante") + guides(fill = guide_legend(title = "Genus"))


# wegen(s)
w %>% filter(cxn %in% c("wegen", "wegens") & position=="post") %>%
  filter(Gender %in% c("f", "m", "n")) %>%
  qbarplot2(cxn, Gender,
                                                            ylab = "Relative Frequency", xlab = "Variant") + guides(fill = guide_legend(title = "Gender"))

# for types rather than tokens
# wegen_s <- filter(w, cxn %in% c("wegen", "wegens") & position=="post")
# wegen_s[!duplicated(wegen_s$lemma),] %>%
#     filter(Gender %in% c("f", "m", "n")) %>%
#   qbarplot2(cxn, Gender,
#                                                             ylab = "Relative Frequenz (Types)", xlab = "Variante") + guides(fill = guide_legend(title = "Genus"))


# only singulars - distinction weak masculines vs. others ---

# add Genus column with distinction between weak and strong nouns
w$Genus <- ifelse(w$Gender=="m" & w$class=="weak", "Masc. weak", w$Gender)
w$Genus <- ifelse(w$Gender=="m" & w$class!="weak", "Masc. strong/mixed", w$Genus)
w$Genus <- ifelse(w$Genus=="f", "Feminine", w$Genus)
w$Genus <- ifelse(w$Genus=="n", "Neuter", w$Genus)

# willen(s)
w %>% filter(cxn %in% c("willen", "willens") & Number=="sg" & Gender %in% c("m", "f", "n")) %>%
  qbarplot2(cxn, Genus, ylab = "Relative Frequency", xlab = "Variant") + 
  guides(fill = guide_legend(title = "Gender"))
# ggsave("gender_distribution_willens_EN.png")

# same for types rather than tokens
# willen_s <- filter(w, cxn %in% c("willens", "willen"))
# willen_s[!duplicated(willen_s$lemma),] %>% filter(Number=="sg" & Gender%in% c("m", "f", "n")) %>%
#   qbarplot2(cxn, Genus, ylab = "Relative Frequenz (Types)", xlab = "Variante") + 
#   guides(fill = guide_legend(title = "Genus"))

# wegen(s)
w %>% filter(cxn %in% c("wegen", "wegens") & Number=="sg" & Gender %in% c("m", "f", "n") & position=="post") %>%
  qbarplot2(cxn, Genus, ylab = "Relative Frequency", xlab = "Variant") + 
  guides(fill = guide_legend(title = "Gender"))
# ggsave("gender_distribution_wegens_EN.png")

# same for types rather than tokens
wegen_s <- filter(w, cxn %in% c("wegen", "wegens") & position=="post")
wegen_s[!duplicated(wegen_s$lemma),] %>% filter(Number=="sg" & Gender %in% c("m", "f", "n")) %>%   qbarplot2(cxn, Genus, ylab = "Relative Frequency, Types", xlab = "Variant") + 
  guides(fill = guide_legend(title = "Gender"))
# ggsave("gender_distribution_Types_wegens_EN.png")

```


## determiner distribution in wegens

In wegens, we expect less determiners for the prenominal use (assuming that prenominal wegens serves as a cleft-opening element). We see that for postnominal wegen and wegens, the definite articles clearly dominates. We find much less definite articles in prenominal wegen and especially wegens.

```{r determiners, warning = FALSE}

# binary coding for determiners
w$Det <- ifelse(w$Determiner=="none", "with determiner", "without determiner")

# factor coding for determiners
# w$Determiner <- factor(w$Determiner, levels = rev(c("none", levels(factor(w$Determiner))[-which(levels(factor(w$Determiner))=="none"|is.na(levels(factor(w$Determiner))))])))
# 
# w %>% filter(cxn %in% c("wegen", "wegens")) %>% filter(position!="both") %>% qbarplot2(cxn, Det, position, wrap100=T)
# 
# w %>% filter(cxn %in% c("wegen", "wegens")) %>% filter(position!="both" & Determiner!="unclear") %>% qbarplot2(cxn, Determiner, position, wrap100=T, color = F)

```


## Number distribution

In plurals, there's no genitive-s that could move. So if we are actually dealing with "wandering" case morphemes, we would expect plurals to be rarer in the wegens/willens datasets than in the canonical comparison data.


```{r number, warning = FALSE}

w %>% filter(cxn %in% c("willens", "willen")) %>% qbarplot2(cxn, Number)
w %>% filter(cxn %in% c("wegens", "wegen") & position=="post") %>% qbarplot2(cxn, Number)


```

## with or without s?

```{r sless, warning = FALSE, out.width='50%', warning = FALSE}
w$s_less <- as.character(w$s_less)
w$s_less <- ifelse(w$s_less=="y", "without s", w$s_less)
w$s_less <- ifelse(w$s_less=="n", "with s", w$s_less)

w %>% filter(cxn %in% c("willen", "willens")) %>%
  filter(s_less %in% c("with s", "without s")) %>%
  qbarplot2(cxn, s_less, ylab="Relative Frequency", xlab="Variant") +
  guides(fill = guide_legend(title = "Genitive \nwith/\nwithout s")) +
  scale_fill_manual(values = brewer.pal(9, "Blues")[c(4,6,8)]) 
# ggsave("willens_sless_genitives_EN.png")

w %>% filter(cxn %in% c("wegen", "wegens") & position=="post") %>%
  filter(s_less %in% c("with s", "without s")) %>%
  qbarplot2(cxn, s_less, ylab="Relative Frequency", xlab="Variant") +
  guides(fill = guide_legend(title = "Genitive \nwith/\nwithout s")) +
  scale_fill_manual(values = brewer.pal(9, "Blues")[c(4,6,8)]) 
# ggsave("wegens_sless_genitives_EN.png")

```


## Random Forest

```{r forest, out.width='50%', warning = FALSE}

# subset with strong masculines and neuters
s <- filter(w, s_less %in% c("with s", "without s") & case=="gen" & !is.na(Freq))

# convert categorical columns to factors
cols <- c("s_less", "cxn", "monoref", "forum", "abbreviation")
s[cols] <- lapply(s[cols], as.factor)

# add apposition column
s$app <- ifelse(s$NP_ext=="app", "y", "n")
s$app <- factor(s$app)

# add binary proper name column
s$name <- factor(ifelse(s$Proper_name!="n", "y", "n"))

# willen(s)
s2 <- select(s, s_less, cxn, monoref, name, app, Freq, pn_or_acro_or_nonnative_in_s, abbreviation, position)

# use more intuitive names
colnames(s2) <- c("s_less", "Variant", "Monoreference", "Proper_name", "Apposition", "Frequency", "Need_for_conservation_s", "Short_word", "Position")

# replace NAs
s2 <- replace_na(s2, list(Need_for_conservation_s = "n"))
s2$Need_for_conservation_s <- factor(s2$Need_for_conservation_s)


# log frequency
s2$LogFreq <- log(s2$Frequency)

# input for trees
s2a <- filter(s2, Variant %in% c("willens", "willen"))
s2b <- filter(s2, Variant %in% c("wegens", "wegen") & Position=="post")



# willen(s) -------------------------------------

# ctree
stree <- ctree(s_less ~ Variant + Monoreference + Apposition + Proper_name + Short_word + Need_for_conservation_s + LogFreq, data = s2a)

# png("willens_tree_EN.png", width = 10, height = 6, un = "in", res = 300)
plot(stree)
# dev.off()


# random forest
sforest <- cforest(s_less ~ Variant + Monoreference + Apposition + Proper_name + Short_word + Need_for_conservation_s + LogFreq, data = s2a)


# get index of concordance - code
# adapted from Tagliamonte & Baayen 2012
s2.trp = treeresponse(sforest)
s2a$PredFOREST <- sapply(s2.trp, FUN = function(v)return(v[2]))
s2a$s_less_bin <- (s2a$s_less=="without s") + 0
somers2(s2a$PredFOREST, s2a$s_less_bin)


# wegen(s) -------------------------------

# ctree
stree <- ctree(s_less ~ Variant + Monoreference + Apposition + Proper_name + Short_word + Need_for_conservation_s + LogFreq, data = s2b)

# png("wegens_tree_EN.png", width = 7.5, height = 5, un = "in", res = 300)
plot(stree)
# dev.off()


# random forest
sforest <- cforest(s_less ~ Variant + Monoreference + Apposition + Proper_name + Short_word + Need_for_conservation_s + LogFreq, data = s2b)


# get index of concordance - code
# adapted from Tagliamonte & Baayen 2012
s2.trp = treeresponse(sforest)
s2b$PredFOREST <- sapply(s2.trp, FUN = function(v)return(v[2]))
s2b$s_less_bin <- (s2b$s_less=="without s") + 0
somers2(s2b$PredFOREST, s2b$s_less_bin)


```



